---
title: "Covid-19 Case Study"
author: "Alex"
date: "2025-06-10"
output:
  pdf_document: default
  html_document: default
---
# Importing Library and Data
This markdown uses `tidyverse`, `forcats`, `ggplot2` and `usmap` libraries.
Make sure to install!
```{r setup, include=FALSE}
library(tidyverse)
library(forcats)
library(ggplot2)
library(usmap)
```

```{r get_data, message=FALSE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",  
                "time_series_covid19_confirmed_global.csv", 
                "time_series_covid19_deaths_US.csv",  
                "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in, file_names)

US_cases <- read_csv(urls[1])
global_cases <- read_csv(urls[2])
US_deaths <- read_csv(urls[3])
global_deaths <- read_csv(urls[4])
```

# Cleaning Data

Although the tables are hard to read at the moment, I think I will keep them 
separate from each other and pivot/clean when needed for a graph.

# Graphs

## Pivoting Global Tables
```{r pivot_table_global, message=FALSE}
global_deaths_long <- global_deaths %>%
  pivot_longer(
    cols = matches("^\\d{1,2}/\\d{1,2}/\\d{2}$"),
    names_to = "Date",
    values_to = "Deaths"
  ) %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%y"))

global_cases_long <- global_cases %>%
  pivot_longer(
    cols = matches("^\\d{1,2}/\\d{1,2}/\\d{2}$"),
    names_to = "Date",
    values_to = "Cases"
  ) %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%y"))
head(global_cases_long)
```

Pivoting the table like this allows us to see the daily cases per region 


## Grabbing Top 8 COuntries
```{r top_contries, message=FALSE}
top_countries <- global_deaths_long %>%
  group_by(`Country/Region`) %>%
  summarise(Total = sum(Deaths, na.rm = TRUE)) %>%
  top_n(8, Total) %>%
  pull(`Country/Region`)
```

The `8` is arbitrary. It gives a wide variety of countries of different 
regions.

```{r graph_cases_and_death, message=FALSE}
global_cases_long %>%
  filter(`Country/Region` %in% top_countries) %>%
  group_by(`Country/Region`, Date) %>%
  summarise(TotalCases = sum(Cases, na.rm = TRUE)) %>%
  ggplot(aes(x = Date, y = TotalCases, color = `Country/Region`)) +
  scale_color_brewer(palette = "Dark2") +
  geom_line() +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = "COVID Cases Over Time by Top 8 Country") +
  theme_minimal()

global_deaths_long %>%
  filter(`Country/Region` %in% top_countries) %>%
  group_by(`Country/Region`, Date) %>%
  summarise(TotalCases = sum(Deaths, na.rm = TRUE)) %>%
  ggplot(aes(x = Date, y = TotalCases, color = `Country/Region`)) +
  scale_color_brewer(palette = "Dark2") +
  geom_line() +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = "COVID Death Over Time by Top 8 Country") +
  theme_minimal()
```

Most countries follow a similar trend. There is also an interesting spike in 
death cases at the beginning of 2022. This could be for multiple reasons like
the holiday season or some major event. But considering that it happened across
the globe, I assume that it was due to people loosening up restrictions.

## Covid-19 in the US

```{r plot_us, message=FALSE}
US_deaths_by_state <- US_deaths %>%
  select(-UID, -iso2, -iso3, -code3, -FIPS, -Lat, -Long_, -Combined_Key, -Admin2) %>%
  pivot_longer(
    cols = matches("^\\d{1,2}/\\d{1,2}/\\d{2}$"),
    names_to = "Date",
    values_to = "Deaths"
  ) %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%y")) %>%
  group_by(Province_State) %>%
  summarise(Max_Deaths = max(Deaths, na.rm = TRUE)) %>%
  mutate(state = Province_State)

plot_usmap(data=US_deaths_by_state, values="Max_Deaths", regions="states") + 
  scale_fill_continuous(
    low = "white", high = "red",
    name = "Max Deaths",
    label = scales::comma
  ) +
  labs(
    title = "Maximum COVID-19 Deaths by U.S. State",
    subtitle = "Cumulative maximum from dataset"
  ) +
  theme(legend.position = "right")
```

From this graph, we can see that the California has the max deaths follow by Florida.
But this graph does not show the population of each state. 

# Model 

For my model, I want to do a basic linear regression to predict tomorrow's cases
based on Today's cases. Since there is so much data, I will be filtering it to
Los Angeles, California. 

```{r create_LA_data, message=FALSE}
losangeles_data <- US_cases %>%
  select(-UID, -iso2, -iso3, -code3, -FIPS, -Lat, -Long_, -Combined_Key) %>%
  pivot_longer(
    cols = matches("^\\d{1,2}/\\d{1,2}/\\d{2}$"),
    names_to = "Date",
    values_to = "Cases"
  ) %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%y")) %>%
  filter(Admin2 == "Los Angeles", Province_State=='California') %>%
  arrange(Date)
losangeles_data
```

Next I want to create a lag and model.

```{r create_LA_lag, message=FALSE}
losangeles_data <- losangeles_data %>%
  mutate(
    daily_cases = Cases - lag(Cases),
    daily_yesterday = lag(daily_cases),
    daily_tomorrow = lead(daily_cases)
  ) %>%
  filter(!is.na(daily_cases), !is.na(daily_yesterday), !is.na(daily_tomorrow))
```

```{r create_lr_model, message=FALSE}
model <- lm(daily_tomorrow ~ daily_cases + daily_yesterday, data = losangeles_data)
summary(model)
```

```{r prediction_and_plotting, message=FALSE}
losangeles_data <- losangeles_data %>%
  mutate(predicted = predict(model, newdata = losangeles_data))

losangeles_data <- losangeles_data %>%
  mutate(
    actual_smoothed = stats::filter(daily_tomorrow, rep(1/7, 7), sides = 1),
    predicted_smoothed = stats::filter(predicted, rep(1/7, 7), sides = 1)
  )

ggplot(losangeles_data %>%
  filter(Date >= as.Date("2021-11-01") & Date <= as.Date("2022-03-01")),
  aes(x = Date)) +
  geom_line(aes(y = actual_smoothed, color = "Actual (7-day avg)"), linewidth = 1) +
  geom_line(aes(y = predicted_smoothed, color="Predicted (7-day avg)"), linetype="dashed", linewidth=1) +
  scale_color_manual(values = c("Actual (7-day avg)" = "blue", "Predicted (7-day avg)" = "red")) +
  labs(
    title = "Zoomed View: LA COVID Surge (Winter 2021–22)",
    y = "7-Day Average of Daily Cases", x = "Date", color = "Legend"
  ) +
  theme_minimal()

```

After a couple of different graphs, I decided to settle with doing a 7-day
rolling average with the graph only showing one of the multiple spikes in Covid-19
cases. We can see here that the prediction does a pretty good job at predict the 
amount of cases.

# Conclusion

Throughout this report, I identified the different tables that were imported
and pivoted those tables to a more readable format. Using the new tables, I was
able to create graphs to show Covid cases and deaths over time as well as a map
of total Covid cases in the US. I also fitted a linear regression model 

## Bias

A big bias is reporting bias. Some countries may under report due to limited 
testing or political pressure. Some people might not even go out to test when
they get Covid but its more likely that Covid death would be reported. Covid 
testing was also unreliable for a long time early on. There would be a lot of
false negative.